---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

```{r echo=FALSE, results="hide", message=FALSE}
#library("txtplot")
library("badger")
```

# rptR

![Build Status](https://travis-ci.org/mastoffel/rptR.svg?branch=master)
[![CRAN total downloads](http://cranlogs.r-pkg.org/badges/grand-total/rptR?color=blue)](https://cran.r-project.org/package=rptR) 
[![Monthly downloads badge](https://cranlogs.r-pkg.org/badges/last-month/rptR?color=blue)](https://CRAN.R-project.org/package=rptR)
[![Codecov test coverage](https://codecov.io/gh/mastoffel/rptR/branch/master/graph/badge.svg)](https://codecov.io/gh/mastoffel/rptR?branch=master)

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

```{r echo=FALSE, results="hide", message=FALSE}
#library("txtplot")
library("badger")
```


### Goal 

The goal of `rptR` is to provide point estimates, confidence intervals and significance tests of the **repeatability** (intra-class correlation coefficient) of measurements based on generalised linear mixed models (GLMMs). The function `?summary.rpt` produces summaries in a detailed format, whereby `?plot.rpt` plots the distributions of bootstrap or permutation test estimates.

### Citation

When using `rptR`, please cite our [paper](https://besjournals.onlinelibrary.wiley.com/doi/10.1111/2041-210X.12797):

Stoffel, M. A., Nakagawa, S., & Schielzeth, H. (2017). rptR: Repeatability estimation and variance decomposition by generalized linear mixed-effects models. *Methods in Ecology and Evolution*, **8(11)**, 1639-1644. `r badge_doi("https://doi.org/10.1111/2041-210X.12797", "green")`

### Installation

You can install the stable version of `rptR`` from CRAN with:

```{r, eval = FALSE}
install.packages("rptR")
```

Or the development version from GitHub with:

```{r gh-installation, eval = FALSE}
# install.packages("remotes")
remotes::install_github("mastoffel/rptR", build_vignettes = TRUE, dependencies = TRUE) 
# manual
browseVignettes("rptR")
```

### Example

```{r example, results = 'hide', warning=FALSE, message=FALSE}
library(rptR)
data(BeetlesBody)

rpts <- rpt(BodyL ~ Treatment + Sex + (1 | Container) + (1 | Population), 
            grname = c("Container", "Population"), data = BeetlesBody, 
            datatype = "Gaussian", nboot = 100, npermut = 100)
```

```{r, results = TRUE, warning=FALSE, message=FALSE}
summary(rpts)
```

#### Plotting

```{r gaussian, tidy = TRUE, fig.show='hold', fig.width=4.2, fig.height=2}
plot(rpts, grname="Container", type="boot", cex.main=0.8, col = "#ECEFF4")
plot(rpts, grname="Population", type="boot", cex.main=0.8, col = "#ECEFF4")
```
    


